---
title: "Airport effects on U.S. County Unemployment Rates"
author: "Robby Powell"
date: "`r format(Sys.Date(), '%Y-%B-%d')`"
output: 
  html_document: 
    keep_md: yes
    toc: yes
---

# Introduction
This is my response to Ari Lamstein's 
R Shapefile [contest](http://www.arilamstein.com/blog/2016/07/12/announcing-r-shapefile-contest/). 

## Analysis Question
My Analysis consists of my efforts to answer the following questions:

  1. What is the effect of distance to the nearest airport on a county's unemployment rate?
  
  2. How does the traffic of the nearest airport affect unemployment rates?
  
## Analysis Plan
The Analysis plan is delineated below. 

1. Import and Clean Data.
2. Visualize each of the main data elements by year.
3. Visualize relationships between variables.
4. Visualize the airport locations on a U.S. Map with County boundaries.
5. Calculate Distances between county centroids and:
    a. the nearest airport
    b. the nearest airport in the Top 50 of average passenger movements
6. Visualize the distances in two maps.
7. Create a regression model and perform diagnostics.

## Data Sources
One of the rules of the contest is that the data are available for everyone to use. In order to meet this requirement, there are links for the data and shapefile locations below. 

  1. Airport Traffic: [Federal Aviation Administration (FAA)](http://www.faa.gov/airports/planning_capacity/passenger_allcargo_stats/passenger/)

  2. Airport Locations: [openflights.org](http://openflights.org/data.html)
  
  3. County Shapefile: [US Census Bureau](https://www.census.gov/geo/maps-data/data/tiger-line.html)
  
  4. Unemployment data: [Bureau of Labor Statistics (BLS)](http://download.bls.gov/pub/time.series/la/)
  
Where feasible, I have also included a series of scripts that may be used to download the exact files that I used for the analysis. These are located in the Appendicies along with the full analysis script for ease of implementation/testing by other analysts.
  
If it is not listed in this section, then it will be listed as either an R package, or it will be something that I create during the analysis and write-up processes.

# Setup
First, we need to set up our script for the analysis. In this case, we are setting up the main directory structure and the packages required for the analysis.

## Directory setup

In my case, I prefer to start the document off with setting up my main directories in variables. This allows me to use the `paste` and `paste0` functions in base R in order to ease the transfer of the scripts to other directories. 

If you change directories, then all you need to do is to alter the `main.dir` variable, and everything else will follow. This reduces the need for changing the directory in multiple locations throughout an analysis.

```{r directorySetup}
# Main directory
main.dir <- "~/spatial-analysis-in-r/"

# Script directory
script.dir <- paste0(main.dir, "scripts/")

# data directory
dat.dir <- paste0(main.dir, "data/")
```

## Package setup

The packages utilized in this analysis are below. Unless otherwise noted, they are available from CRAN.

The comments after the packages explain why I am using them.

```{r loadPackages, warning = FALSE, error = FALSE, message = FALSE}
# data import
library(openxlsx)

# data cleaning
library(tidyr)
library(dplyr)
library(magrittr)
library(stringr)

# data table display
library(Hmisc)
library(tables)

# general plotting
library(ggplot2)
library(gridExtra)

# refine plot displays
library(scales)
library(RColorBrewer)

# mapping
library(rgdal)
library(GISTools)
```

# Data Import and Cleaning
In this section, I will import and clean the data.

## BLS Data
First, I will import the main data set, whihc is labeled `la.data.64.County`. This contains all of the unemployment numbers.

```{r blsDataPrep1}
# Import UE Data
fname <- paste0(dat.dir, "/bls/la.data.64.County")
bls.dat <- read.delim(fname, header = TRUE, stringsAsFactors = FALSE, 
                      sep = "\t")

# trim white space in series_id
bls.dat$series_id <- trimws(bls.dat$series_id)

# change value to numeric
bls.dat$value <- as.numeric(bls.dat$value)

head(bls.dat)
```

Upon examination, the missing values are for the years 2005 and 2006, which means that they are not of interest to this analysis.

After getting the basic preparation for the main dataset, it is time to import the crosswalks for easier identification.

```{r blsDataPrep2}
# Import la.measure
fname <- paste0(dat.dir, "/bls/la.measure")
bls.measure <- read.delim(fname, header = TRUE, stringsAsFactors = FALSE, 
                      sep = "\t")
```

After looking at this data, the variables are mis-labeled. These will be fixed below.

```{r blsDataPrep3}
# fix variables
bls.measure$measure_text <- bls.measure$measure_code
bls.measure$measure_code <- as.numeric(rownames(bls.measure))

bls.measure
```

Import and examine the period definitions.
```{r blsDataPrep4}
# la.period
fname <- paste0(dat.dir, "/bls/la.period")
bls.period <- read.delim(fname, header = TRUE, stringsAsFactors = FALSE, 
                      sep = "\t")

bls.period
```

Import series definitions, which contains the names of the counties.
```{r blsDataPrep6}
# la.period
fname <- paste0(dat.dir, "/bls/la.series")
bls.series <- read.table(fname, header = TRUE, stringsAsFactors = FALSE, 
                      sep = "\t")

# trim white space from series_id
bls.series$series_id <- trimws(bls.series$series_id)

head(bls.series)
```

Now that all of the necessary data elements are imported, it's time to do some cleaning on the BLS dataset. 

In this case, the data needs to be joined with the proper identifiers, and then the dataset will need to be reduced.

Because we are interested in the annual averages, we will filter for period `M13`. In addition, we will filter for the years 2011-2015. After we have joined the data, we will select only required variables.

```{r blsDataPrep7}
# filter for period M13
bls.dat <- bls.dat[bls.dat$period == "M13",]

# filter for year
bls.dat <- bls.dat[bls.dat$year %in% c(2011:2015),]
 
# Join data together via a left join.
bls.dat <- merge(bls.dat, bls.series,
                 by.x = "series_id", by.y = "series_id",
                 all.x = TRUE)

bls.dat <- merge(bls.dat, bls.measure,
                 by.x = "measure_code", by.y = "measure_code", all.x = TRUE)

# filter for unemployment rate
bls.dat <- bls.dat[bls.dat$measure_text == "unemployment rate", ]

# reduce dataset
bls.dat <- bls.dat[, c("series_id", "year", "value", 
                       "series_title", "measure_text")]

# Create county name and state variables
bls.dat$county <- str_extract(bls.dat$series_title, 
                              "[A-za-z.//]*\\s*[A-za-z.//]*\\s*[A-Za-z.//]*,")
bls.dat$county <- trimws(gsub(",", "", bls.dat$county))

bls.dat$state <-  str_extract(bls.dat$series_title, "[A-Z]{2}")

# Extract FIPS Code
bls.dat$FIPS <- substr(bls.dat$series_id, 6, 10)

# Select columns
bls <- bls.dat[c("FIPS", "county", "state", "year", "value")]

# Change names
names(bls) <- c("FIPS", "county", "state", "year", "UE")
```

Here is the final BLS dataset.

```{r blsDataPrep8}
head(bls)
summary(bls)
```


## Airport Data
Here, we import and check the airport dataset. This dataset is really clean, so it will be utilized as-is.

```{r airportPrep}
fname <- paste0(dat.dir, "airports/airports.dat.txt")
airports <- read.csv(fname, header = TRUE, stringsAsFactors = FALSE)

summary(airports)
head(airports)
```

## FAA Airport Traffic Data
The FAA Airport Data consists of three Excel spreadsheets. These will be imported and combined before they are cleaned.

```{r faaPrep}
fname <- paste0(dat.dir, "faa/CY12AllEnplanements.xlsx")
faa.cy12 <- openxlsx::read.xlsx(fname)

fname <- paste0(dat.dir, "faa/cy13-all-enplanements.xlsx")
faa.cy13 <- openxlsx::read.xlsx(fname)

fname <- paste0(dat.dir, "faa/CY14-all-enplanements.xlsx")
faa.cy14 <- openxlsx::read.xlsx(fname, sheet = "data")

fname <- paste0(dat.dir, "faa/preliminary-cy15-all-enplanements.xlsx")
faa.cy15 <- openxlsx::read.xlsx(fname)

# Select Annual Data, Rename columns, and add year.
faa.cy11 <- faa.cy12[, c("ST", "Locid", "City", "Airport.Name", 
                         "CY.11.Enplanements")]
names(faa.cy11) <- c("state", "id", "city", "airport.name", "enplanements")
faa.cy11$year <- 2011

faa.cy12 <- faa.cy12[, c("ST", "Locid", "City", "Airport.Name", 
                         "CY.12.Enplanements")]
names(faa.cy12) <- c("state", "id", "city", "airport.name", "enplanements")
faa.cy12$year <- 2012

faa.cy13 <- faa.cy13[, c("ST", "Locid", "City", "Airport.Name",
                         "CY.13.Enplanements")]
names(faa.cy13) <- c("state", "id", "city", "airport.name", "enplanements")
faa.cy13$year <- 2013

faa.cy14 <- faa.cy14[, c("ST", "Locid", "City", "Airport.Name",
                         "CY.14.Enplanements")]
names(faa.cy14) <- c("state", "id", "city", "airport.name", "enplanements")
faa.cy14$year <- 2014

faa.cy15 <- faa.cy15[, c("ST", "Locid", "City", "Airport.Name",
                         "CY.15.Enplanements")]
names(faa.cy15) <- c("state", "id", "city", "airport.name", "enplanements")
faa.cy15$year <- 2015

# Combine
faa <- rbind(faa.cy11, faa.cy12, faa.cy13, faa.cy14, faa.cy15)

summary(faa)
head(faa)
```



## Shapefile Import

## Identify Counties associated with Airports

## Calculate County Distance to the Nearest Airport

## Create Final Data Set

# Exploratory Analysis

# Setting up the Model Data Set

# Relationship between distance to airport and unemployment rate

## Training and Testing Datasets

## Training the Model

## Testing the Model

## Analyzing the Model

# Discussion

# Conclusions

# Computer Environment
The computer used for the analysis was a Late 2013 15" MacBook Pro, with 16 GB 1600 MHz DDR3 RAM and a 2.6 GHz Intel Core i7 Processor.

```{r computerInfo}
sessionInfo()
```

# Appendices
This section will show the full code of the analysis scripts.

## Appendix A: Complete Analysis
```{r appendixA, eval = FALSE}


```

## Appendix B: Airport Location from openflight.org
```{r appendixB, eval = FALSE}
# File url
file <- "https://raw.githubusercontent.com/jpatokal/openflights/master/data/airports.dat"

# destination
dest <- "~/spatial-analysis-in-r/data/airports/airports.dat.txt"

# download file
download.file(file, dest)

# Check downloaded file
airports <- read.csv(dest, header = FALSE)

# name columns
names(airports) <- c("airport.id", "name", "city", "country", "iata.faa.code",
                     "icao.code", "latitude", "longitude", "altitude",
                     "UTC.offset", "DST", "time.zone.olson")

# save file with column names
  write.csv(airports, dest, row.names = FALSE)
```

## Appendix C: County Shapefile from Census Bureau
```{r appendixC, eval = FALSE}
# County shapefile url
url <- "ftp://ftp2.census.gov/geo/tiger/TIGER2015/COUNTY/tl_2015_us_county.zip"

# path
path <- "~/spatial-analysis-in-r/data/census/"

# destination
dest <- paste0(path, "tl_2015_us_county.zip")

# download file
download.file(url, dest)

# unzip file
unzip(dest, exdir = path, junkpaths = TRUE)
```

## Appendix D: BLS Data
```{r appendixD, eval = FALSE}

# Set main part of url for downloading the files
main.url <- "http://download.bls.gov/pub/time.series/la/"

# Set main directory for download
download.dir <- "~/spatial-analysis-in-r/data/bls/"

# List of files to download
file.list <- c("la.area", "la.area_type", "la.data.64.County", "la.measure",
               "la.period", "la.series", "la.state_region_division")

# combine download urls
download.urls <- unlist(lapply(main.url, FUN = "paste0", file.list))

# combine for file locations
download.locations <- unlist(lapply(download.dir, FUN = "paste0", file.list))

# download files
for (file in 1:length(file.list)) {
  if (!(file.list[file] %in% dir(download.dir))) {
    download.file(url = download.urls[file], destfile = download.locations[file])
  }
}
```